# PGRNseq Sample and Phenotype Processing

```{r load_packages, include=FALSE}
.libPaths('~/R/x86_64-pc-linux-gnu-library/3.0/')
library(ggplot2)
library(mclust)
library(SNPRelate)
```

```{r define_constants, include=FALSE}
PHENO.FN <- '../data/rdata/pheno.Rdata' # dataframe of processed phenotype data
exome.fbase <- '../data/exomechip_data/innocenti_082613'
```

```{r define_functions, include=FALSE}
## weird sample ID format for italian samples
removeACO <- function(x) {
  return(gsub('\\+AC0', '', x))
}
## formatting of dosages
removemgm2 <- function(x) {
  return(gsub(' mg\\/m2' ,'', x))
}
## correct inconsistency in Erasmus sample ID encoding
homogenizeSampleIds <- function(x) {
  return(gsub('M0', 'MO', x))
}
```

```{r load_phenotye, include=FALSE}
## read in phenotype (ANC) and covariate data
anc <- read.table('../data/annotation/ANC_Nadir_FULL_061614.csv',
                  sep='\t',
                  header=T,
                  stringsAsFactors=F)
## clean up the italian sample IDs
anc$Sample <- unlist(lapply(anc$Subject_ID, removeACO) )
anc$Sample <- unlist(lapply(anc$Sample, homogenizeSampleIds) )
covars <- read.table('../data/annotation/Covariates_Vasa_newIDs.csv',
                     head=T,
                     sep='\t',
                     stringsAsFactors=F)
covars$SM_ID <- as.character(covars$SM_ID)
clean.pheno <- merge(anc, covars, by.x="Sample", by.y="Sample.ID")
nomiss.pheno <- clean.pheno[which(!is.na(clean.pheno$ANC_nadir_1000cells.per.mcL) &
                                  !is.na(clean.pheno$SM_ID) ),]

## make factor variables out of the covariates
nomiss.pheno$Study.site <- as.factor(nomiss.pheno$Study.site)
## homogenize the sex encodings
nomiss.pheno$Sex[which(nomiss.pheno$Sex=='Male')] <- '1'
nomiss.pheno$Sex[which(nomiss.pheno$Sex=='M')] <- '1'
nomiss.pheno$Sex[which(nomiss.pheno$Sex=='Female')] <- '0'
nomiss.pheno$Sex[which(nomiss.pheno$Sex=='F')] <- '0'
nomiss.pheno$Race[which(nomiss.pheno$Race == "")] <- NA
## clean up the dose variables
nomiss.pheno$Dose..mg. <- as.numeric(unlist(lapply(list(nomiss.pheno$Dose..mg.), removemgm2)))
```

Format the phenotype data for downstream processing.

```{r format_pheno}
pheno.df <- data.frame(id=nomiss.pheno$SM_ID,
                       sex=as.factor(nomiss.pheno$Sex),
                       ANC=as.numeric(nomiss.pheno$ANC_nadir_1000cells.per.mcL),
                       site=as.factor(nomiss.pheno$Study.site),
                       race=as.factor(nomiss.pheno$Race),
                       dose=as.factor(nomiss.pheno$Dose..mg.))
```


## Expirical Distributions of Phenotypes
```{r anc_plot}
anc.plt <- ggplot(pheno.df, aes(x=ANC)) + geom_histogram()
print(anc.plt + ggtitle('ANC'))
```

```{r loganc_plt}
loganc.plt <- ggplot(pheno.df, aes(x=log10(ANC))) + geom_histogram()
print(loganc.plt + ggtitle('log10(ANC)'))
```


```{r unlink_existing, include=FALSE}
## cleanup SNPRelate files if they exist
if(!file.access(paste0(exome.fbase, '.gds'))) unlink(paste0(exome.fbase, '.gds'))
## load exome chip data from PLINK files
snpgdsBED2GDS(bed.fn=paste0(exome.fbase, '.bed'),
              bim.fn=paste0(exome.fbase, '.bim'),
              fam.fn=paste0(exome.fbase, '.fam'),
              out.gdsfn=paste0(exome.fbase, '.gds'))
exome.geno <- openfn.gds(filename=paste0(exome.fbase, '.gds'))
```

## PCA
Prune SNP set for LD, filter on MAF, and run PCA.
```{r pca}
ld.snpset <- snpgdsLDpruning(exome.geno)
pca <- snpgdsPCA(exome.geno, snp.id = unlist(ld.snpset), maf = 0.05)
pca.df <- data.frame(sample.id=pca$sample.id, eigenvect=pca$eigenvect)
```

Look at the first two PCs, and color the samples by self-reported ancestry.
```{r reported_plot, echo=FALSE}
reported.plt <- ggplot(pheno.df, aes(x=eigenvect.1, y=eigenvect.2, color=factor(race))) +
                      geom_point()
print(reported.plt + ggtitle('PCA w/ self reported ancestry'))
```

## Model-based Clustering
Add principal components to the phenotype data frame, then cluster samples in the first two PCs.
```{r cluster}
pheno.df <- merge(pca.df, pheno.df, by.x = "sample.id", by.y = "id")
names(pheno.df)[which(names(pheno.df) == "sample.id")] <- "id"
cluster <- Mclust(pheno.df[,c("eigenvect.1", "eigenvect.2")])
pheno.df$cluster <- cluster$classification
```

Examine the results of our PCA and clustering the first two PCs.
```{r cluster_plot, echo=FALSE}
cluster.plt <- ggplot(pheno.df, aes(x=eigenvect.1, y=eigenvect.2, color=factor(cluster))) +
                      geom_point()
print(cluster.plt + ggtitle('PCA w/ model-based clustering'))
```

Taking a closer look at the clustering of the Europeans. We may be able to explain some genetic variation just by site of ascertainment.
```{r euro_zoom_plot, echo=FALSE}
euro.zoom.plt <- ggplot(pheno.df, aes(x=eigenvect.1, y=eigenvect.2, color=factor(site))) +
                        geom_point() +
                        xlim(-0.05, 0.025) + ylim(-0.05, 0.05)
print(euro.zoom.plt + ggtitle('PCA: european cluster colored by site'))
```

```{r save, include=FALSE}
## dump phenotype data to disk
save(pheno.df, file=PHENO.FN)
```
